Modelling fine-scale correlation structures using hidden Markov models

### 2020-07-16 Report/Sections/intro.tex
Updates from Overleaf

The field of animal movement is in the midst of a ``data renaissance" where advancements in tagging technology have given rise to an explosion of data available for statistical modeling. In particular, tagging technologies are capable of recording observations at rates of tens of hertz, resulting in time series containing millions of observations over the course of several hours. In response, researchers have introduced a variety of new statistical techniques to infer animal behavior from movement data \citep{Hooten:2017}.

One of the most prevalent techniques in the literature of late is the hidden Markov model (HMM), where observations depend upon the state of an associated unobserved behavioral process following Markovian dynamics \citep{Patterson:2017}. Importantly, under the traditional HMM model, subsequent observations are assumed to be independent from one another conditioned on the underlying behavioral process. However, this assumption is often violated in the real world processes, especially when observations are taken a high frequencies. For example, the location of an animal at a given time is highly correlated with the location of that animal one second later. Several publications have dealt with this issue in the past, including the hidden movement Markov model (HMMM) \citep{Whoriskey:2016} and the conditionally auto-regressive hidden Markov model (CarHMM) \citep{Lawler:2019}. The CarHMM in particular explicitly models auto-correlation into an HMM while maintaining the structure needed to run the forward algorithm. It also only adds one additional parameter per possible hidden state.

Another issue that arises in high-frequency data is that several simultaneous behavioral processes may occur at different time scales. In this work we consider the example where killer whales perform a variety of different types of dives at a coarse scale, but also exhibit many different types of swimming behaviors within dives at a fine scale. One solution to this issue is to use a hierarchical hidden Markov model (HHMM) \citep{Barajas:2017} \citep{Adam:2019}. HHMMs model the entire time series in question as a nested structure of hidden Markov models (HMM) where each HMM corresponds to one behavioral process.

At the shortest time scales, however, observations often exhibit complicated dependence structures which cannot be easily captured by traditional HMMs, CarHMMs, or HHMMs. Examples included periodic fluking behavior in killer whales off the coast of Vancouver, BC, and swimming patterns of horn sharks of the coast of Southern California \citep{Adam:2019}. One solution is to model the fine-scale behavior using a continuous-time model, which usually involves modelling the dynamics of an animal as the solution to a stochastic differential equation. Continuous-time models are more flexible than their discrete-time counterparts and can incorporate observations taken at irregular time intervals. However, they are often computationally intractable and require approximate inference techniques such as Markov-chain Monte Carlo (MCMC) methods to perform inference.

For periodic behavior in particular, One way to avoid the use of continuous time models is to use signal processing techniques such as the Fourier transform on the raw data. The advantages of using Fourier analysis within an HMM has been recently demonstrated in the context of describing daily behavioral cycles of marine mammals \citep{Heerah:2017}. In addition, Fourier analysis has previously been used in the field of animal movement to explain animal behavior \citep{Fehlmann:2017} and specifically fluking \citep{Shorter:2017} from accelerometer data. Thus, incorporating Fourier analysis of accelerometer data within the structure of an HMM appeared a promising simple approach to account for additional correlation in data that is cyclical in nature.

We consider several models- the classical hierarchical hidden Markov model (HHMM) \citep{Barajas:2017}, the CarHMM model of \citep{Lawler:2019}, and our new model that blends the two (CarHHMM). The HHMM consists of a coarse-scale process and a fine-scale process, with standard HMMs modelling both the coarse- and fine-scale processes. The CarHMM modifies the usual HMM by modelling additional dependence into the sequence of observed data. In our combination of the HHMM and the CarHHMM models, we keep the usual Markov structure for the coarse-scale process, but use the CarHMM model for the fine-scale process.

>>>>

The field of animal movement is in the midst of a ``data renaissance" where advancements in tagging technology have given rise to an explosion of data available for statistical modeling. In particular, some tagging technologies are capable of recording observations at rates of tens of hertz, resulting in time series containing millions of observations over the course of several hours. In response, researchers have introduced a variety of new statistical techniques to infer animal behavior from movement data \citep{Hooten:2017}.

One of the most prevalent techniques in recent literature is the hidden Markov model (HMM), where observations depend upon the state of an associated unobserved behavioral process following Markovian dynamics \citep{Patterson:2017}. Importantly, under the traditional HMM model, subsequent observations are assumed to be independent from one another after conditioning on the underlying behavioral process. However, this assumption is often violated in real world processes, especially when observations are taken at high frequencies. For example, the location of an animal at a given time is highly correlated with the location of that animal one second later. Several publications have dealt with this issue in the past, including the hidden movement Markov model (HMMM) \citep{Whoriskey:2016} and the conditionally auto-regressive hidden Markov model (CarHMM) \citep{Lawler:2019}. The CarHMM in particular explicitly models auto-correlation into an HMM while maintaining the structure needed to run the forward algorithm. It also only adds one additional parameter per possible hidden state.

Another issue that arises in high-frequency data is that several simultaneous behavioral processes may occur at different time scales. In this work, we consider an example where killer whales exhibit a variety of different types of dives at a coarse scale, but also exhibit many different types of swimming behaviors within dives at a fine scale. One solution to this issue is to use a hierarchical hidden Markov model (HHMM) \citep{Barajas:2017} \citep{Adam:2019}. HHMMs model the entire time series in question as a nested structure of hidden Markov models (HMM) where each HMM corresponds to one behavioral process.

All HMMs presented so far assume Markovian dynamics in the underlying process (i.e. that any observation $Y_t$ depends only on the behavioral state $X_t$ and $Y_{t-1}$ when conditioned on all previous time steps). At the shortest time scales, however, observations often exhibit complicated dependence structures which cannot be easily captured by traditional HMMs, CarHMMs, or HHMMs. Examples included periodic fluking behavior in killer whales off the coast of Vancouver, BC, and swimming patterns of horn sharks of the coast of Southern California \citep{Adam:2019}.  With improvements in tagging technology allowing for data to be collected at very high frequencies, noisy and non-Markovian fine-scale processes are likely to persist.

One solution is to model the fine-scale behavior using a continuous-time model, which involves modelling the dynamics of an animal as the solution to a stochastic differential equation. Continuous-time models are more flexible than their discrete-time counterparts and can incorporate observations taken at irregular time intervals. Moreover, they are often computationally intractable and require approximate inference techniques such as Markov-chain Monte Carlo (MCMC) methods to perform inference.

For periodic behavior in particular, one way to avoid the use of continuous time models is to use signal processing techniques such as the Fourier transform on the raw data. The advantages of using Fourier analysis within an HMM has been recently demonstrated in the context of describing daily behavioral cycles of marine mammals \citep{Heerah:2017}. In addition, Fourier analysis has previously been used in the field of animal movement to explain animal behavior \citep{Fehlmann:2017}, specifically fluking \citep{Shorter:2017} from accelerometer data. Thus, incorporating Fourier analysis of accelerometer data within the structure of an HMM appeared a promising simple approach to account for additional correlation in data that is cyclical in nature.

We consider several models: the classical hierarchical hidden Markov model (HHMM) \citep{Barajas:2017}, the CarHMM model of \citep{Lawler:2019}, and our new model that blends the two (CarHHMM). The HHMM consists of a coarse-scale process and a fine-scale process, with standard HMMs modelling both the coarse- and fine-scale processes. The CarHMM modifies the usual HMM by modelling additional dependence into the sequence of observed data. In our combination of the HHMM and the CarHHMM models, we keep the usual Markov structure for the coarse-scale process, but use the CarHMM model for the fine-scale process.

### 2020-07-16 Report/Sections/models.tex
Updates from Overleaf

In this section we first remind the reader of the definition of a hidden Markov model (HMM). We then review three variations on the HMM; the CarHMM, HHMM, and HMM-DFT. We also briefly describe a continuous-time model and its connection with the CarHMM. Finally, we generalize the hierarchical structure of these models and describe how they can be flexibly altered and combined to form new models.

@@

A \textit{hidden Markov model}  is comprised of a sequence of  unobserved states $X_t$, $t = 1, \ldots, T$, and an associated sequence of  possibly high-dimensional observations $Y_t$, $t = 1, \ldots, T$,
The $Y_t$s are often referred to as ``emissions'' and the index $t$ typically refers to time. 
The $X_t$s form a Markov chain and take possible values $1, \ldots, N$. Their distribution is governed by the distribution of the initial state $X_1$ and the $N \times N$ transition probability matrix $\Gamma$ where $\Gamma_{ij} = \Pr(X_{t+1} = j | X_t = i)$, for $t=1,\ldots, T-1$, and $i, j = 1,\ldots, N$. 
%
We assume that $X_1$ follows the chain's stationary distribution, which is denoted by $\delta \in \bbR^N$, with $i$th component
$\delta_i = \Pr\{X_1 = i\},~ i = 1,\ldots,N.$
%
Recall that a Markov chain's stationary distribution is determined by its probability transition matrix via $\delta = \delta \Gamma$ and $\sum_{i=1}^N \delta_i = 1$.
%
The distribution of an emission $Y_t$ depends only on the corresponding state $X_t$ and no other observations or hidden states: $p\left(y_t|\{X_1,\ldots, X_T\},\{Y_1,\ldots, Y_T\}/ \{Y_t\}\right) = p(y_t|X_t)$.
%
These conditional distributions are governed by state-dependent parameters: if $X_t = i$, the state-dependent parameter is $\theta^{(i)}$ and we denote the conditional distribution of $Y_t$ given $X_t=i$ by its conditional density or probability mass function, denoted $f^{(i)}(\cdot ; \theta^{(i)})$, or sometimes  $f^{(i)}(\cdot)$.
%
(Fig \ref{fig:HMM}) represents the dependence structure of an HMM.

@@

%
We denote the conditional distribution of $Y_t$ given $Y_{t-1}= y_{t-1}$ and $ X_t=i$ as $f^{(i)}( \cdot | y_{t-1}; \theta^{(i)})$ or simply $f^{(i)}( \cdot | y_{t-1})$.
As an example, we could assume that this conditional distribution is normal with parameters $\theta^{(i)} = \{\mu^{(i)},\sigma^{(i)},\phi^{(i)}\}$ where:
%
\[
\mathbb{E}(Y_t|Y_{t-1} = y_{t-1},X_t=i) = \phi^{(i)} ~ y_{t-1} ~+ ~(1-\phi^{(i)})  ~\mu^{(i)}
\]
and
\[
\mathbb{V}(Y_t| Y_{t-1} =y_{t-1}, X_t = i) = (\sigma^{(i)})^2.
\]
%
The likelihood for the CarHMM can be easily calculated using the forward algorithm. As previously, let $y$ be the vector of observed emissions. Then
\begin{equation}
    \calL_{\text{CarHMM}}(y;\Theta,\Gamma) = \delta \prod_{t=2}^T \Gamma P(y_t|y_{t-1};\Theta) \mathbf{1}_N
    \label{CarHMM_likelihood}
\end{equation}
where
%
$P(y_t|y_{t-1};\Theta)$ is an $N \times N$ diagonal matrix with $ii$th entry equal to $f^{(i)}(y_t|y_{t-1}; \theta^{(i)})$.
%
\subsection{Continuous-time processes}

While CarHMMs can model auto-correlation within the observation sequence, they fail when observations are taken at irregular time intervals. Another way to model a stochastic process is to use a stochastic differential equation, or SDE. As an example, \citep{Michelot:2019} model the movement of an animal as the solution to the following stochastic differential equation:
%
\begin{equation}
    \label{eqn:SDE}
    dY_t = \beta^{(X_t)}(\gamma^{(X_t)} - Y_t)dt + \omega^{(X_t)} dW_t
\end{equation}
%
where $X_t$ follows some stochastic process which defines the hidden behaviour of the animal at time $t$, $\beta^{(X_t)}$ relates to rate at which the process returns to its mean value, $\gamma^{(X_t)}$ is the long-term mean value of the process, $\omega^{(X_t)}$ is related to short-term variance, and $W$ is a Wiener process. This SDE is referred to as a state-switching Ornstein-Uhlenbeck process. Unlike HMMs, the time index $t \in \mathbb{R}$ exists in continuous time and is not necessarily an integer. If the behavioral state $X_t$ is known and does not change between observations, the solution to (eqn \ref{eqn:SDE}) is known to be the following \citep{Michelot:2019}:
%
\begin{align}
    Y_{t+\Delta t} | X_{t} \sim \mathcal{N}\left((1-e^{-\beta^{(X_t)}\Delta t})\gamma^{(X_t)} + e^{-\beta^{(X_t)}\Delta t} Y_t,\quad \frac{\omega^{(X_t)^2}}{2\beta^{(X_t)}} (1-e^{-2\beta^{(X_t)}\Delta t})\right)
    \label{eqn:OU_sol}
\end{align}
%
where $\Delta t$ is the time difference between any two observations $Y_t$ and $Y_{t+\Delta t}$. Most continuous time models are difficult to incorporate into an HHMM and require MCMC methods to fit. However, under certain conditions, the CarHMM is equivalent to a state-switching Ornstein-Uhlenbeck process. This gives new interpretation to the learned parameters of the CarHMM in the context of a continuous-time model.

@@

A \textit{hierarchical hidden Markov model} contains both a coarse-scale process and a fine-scale process, both of which are HMMs. The coarse-scale process is a hidden Markov model as defined previously, where $X_1, \ldots, X_T$ make up an unobserved Markov chain with $N$ possible states and $Y_1,\ldots, Y_T$ are the corresponding observed responses.   
%
In the hierarchical setting, each state $X_t$ emits another sequence of fine-scale unobserved states, $X_t^* \equiv (X_{t,1}^*,\ldots, X_{t,T_t^*})$ and a sequence of fine-scale observed emissions, $Y_t^* \equiv (Y_{t,1}^*,\ldots, Y_{t,T_t^*})$. This fine-scale process can be either instead of or in addition to the coarse-scale observations $Y_t$. The length of each fine-scale process is denoted as $T^*_t$ and can often be controlled by the practitioner. Conditional on the coarse-scale hidden state $X_t$, the joint fine-scale process, ($X_t^*, Y_t^*$), is a hidden Markov model with parameters depending on the value of $X_t$. Specifically, given that $X_t=i$, $X_t^*$ is a Markov chain with states $1,\ldots, N_t^*$. The distribution of $X_t^*$, conditional on $X_t=i$, is given by an $N^*_t \times N^*_t$ transition probability matrix $\Gamma^{*(i)}$ and initial probability, denoted by the $N_t^*$-vector $\delta^{*(i)}$, which we assume is equal to the stationary distribution of the chain. For simplicity, we take $N_t^* \equiv N^*$ although this is not necessary.
%
The distribution of $Y_{t, t^*}$ given $X_{t, t^*}=i^*$ and $X_t=i$ is governed by a parameter $\theta^{(i,i^*)}$, and has density or probability mass function denoted $f^{*(i,i^*)}\left(\cdot; \theta^{(i,i^*)}\right)$ or simply $f^{*(i,i^*)}(\cdot)$. Let $\Theta^{*(i)}=\left(\theta^{(i,1)}, \ldots, \theta^{(i,N^*)}\right)$, the fine-scale emission parameter vector corresponding to $X_t=i$.
%
Given the coarse-scale states, $X_1,\ldots, X_T$, the $T$ fine-scale processes $(X_1^*, Y_1^*), \ldots, (X_T^*, Y_T^*)$, are independent HMMs. Depending upon the process being modeled, it is possible to force certain parameters to be shared across different coarse or fine states. For example, in the killer whale case study in section \ref{sec:case_study} we force the fine-scale emission parameters to be shared across coarse-scale hidden states, i.e. $\theta^{(1,i^*)} = \ldots = \theta^{(N,i^*)}$ for $i^* = 1, \ldots, N^*$. (Fig \ref{fig:HHMM}) shows a graphical representation of the dependence structure for an HHMM.

Due to the nested structure of the hierarchical hidden Markov model, the likelihood remains easy to calculate via the forward algorithm.
%
Let $y$ be the $T$-vector of the observed coarse-scale emissions and
$y^*$ be the $(T_1^* + \cdots T_T^*)$-vector of the observed fine-scale emissions.
%
Let $\Theta^*$ denote the collection of all fine-scale emission parameters,
$\Theta^{*(i)}$, $i=1,\ldots, N$, and let $\Gamma^*$ denote the collection of all fine-scale transition probability matrices, $\Gamma^{*(i)}$, $i=1,\ldots, N$.
%
Then the likelihood of the observed data is
%
\[
\calL_{\text{HHMM}}(y,y^*;\Theta,\Theta^*,\Gamma,\Gamma^*) = \delta P(y_1,y_1^*;\Theta,\Theta^*,\Gamma^*) \prod_{t=2}^T \Gamma P(y_t,y_t^*;\Theta,\Theta^*,\Gamma^*) \mathbf{1}_N
\]
%
where $P(y_t,y_t^*;\Theta,\Theta^*,\Gamma^*)$ is an $N \times N$ diagonal matrix with $ii$th entry corresponding to $X_t=i$ and equal to 
$f^{(i)}(y_t)\calL_{\text{HMM}}\left(y_t^*;\Theta^{*(i)},
\Gamma^{*(i)}\right)$.

@@

The HMM with discrete Fourier transform, or HMM-DFT, incorporates hierarchical structure into an HMM differently than an HHMM. In particular, the fine-scale process is no longer modeled with an HMM and instead summarized using its Fourier transform. For simplicity of presentation, we assume that the length of the fine-scale processes is constant, i.e. that $T^*_t = T^*$, although this need not be the case in general. Suppose that the fine-scale process $y^*_t$ does not switch states hidden states, but does exhibit significant periodic behavior. Then, we suggest using the discrete Fourier transform (DFT) on $y^*_t$:
%
\begin{align*}
    DFT\{y^*_t\}(k) := \hat{y}^{*(k)}_{t} = \sum_{t^* = 1}^{T^*} y^*_{t,t^*}\exp\left(-i \frac{2\pi k}{T^*} (t^*-1)\right) \qquad k = 0, 1, \ldots, T^*-1
\end{align*}
%
Summary statistics can then drastically reduce the dimension of $\hat{y}^*_t$. One such example is as follows:
%
\begin{equation}
    \label{eqn:z}
    z_t^{*(1)} := \mathcal{R}\left(\hat{y}^{(0)}_t\right) \qquad z_t^{*(2)} := \frac{1}{T^*}\sum_{k=1}^{\tilde{\omega}}|\hat{y}^{(k)}_t|^2
\end{equation}
%
$z_t^{*(1)}$ is the average value of $y^*_t$ and $z_t^{*(2)}$ is the squared 2-norm of the component of $y^*_t$ that can be attributed to frequencies between $1$ and $\tilde{\omega}$ periods per window length $T^*$. The maximum frequency $\tilde{\omega}$ is a problem-specific tuning parameter which should be selected with care. These summary statistics are just one possible choice to describe each window; other choices include the dominant frequency and amplitude of $y^*_t$. Figure (\ref{fig:fourier_example}) visually displays the process of transforming $y^*$ into $z^*$.

Once $z^*_t$ is calculated, it can simply be treated as an observation of the coarse-scale HMM and incorporated into the emission distribution $f^{(i)}\left(y_t,z^*_t;\theta^{(i)}\right)$, or more succinctly $f^{(i)}\left(y_t,z^*_t\right)$. In total, the likelihood of the hierarchical HMM-DFT is as follows:
\begin{equation}
    \calL_{\text{HMM-DFT}}(y,z^*;\Theta,\Gamma) = \delta P(y_1,z^*_1;\Theta) \prod_{t=2}^T \Gamma P(y_t,z^*_t;\Theta) \mathbf{1}_N
    \label{HMMDFT_likelihood}
\end{equation}
where $P(y_t,z^*_t;\Theta)$ is an $N \times N$ diagonal matrix with $ii$th entry equal to $f^{(i)}\left(y_t,z^*_t;\theta^{(i)}\right)$.

>>>>

In this section we first remind the reader of the definition of a hidden Markov model (HMM). We then review three variations on the HMM: the CarHMM, HHMM, and HMM-DFT. We also briefly describe a continuous-time model and its connection with the CarHMM. Finally, we generalize the hierarchical structure of these models and describe how they can be flexibly altered and combined to form new models.

@@

A \textit{hidden Markov model}  is comprised of a sequence of  unobserved states $X_t$, $t = 1, \ldots, T$, and an associated sequence of  possibly high-dimensional observations $Y_t$, $t = 1, \ldots, T$,
The $Y_t$'s are often referred to as ``emissions'' and the index $t$ typically refers to time. 
The $X_t$'s form a Markov chain and take possible values $1, \ldots, N$. Their distribution is governed by the distribution of the initial state $X_1$ and the $N \times N$ transition probability matrix $\Gamma$ where $\Gamma_{ij} = \Pr(X_{t+1} = j | X_t = i)$, for $t=1,\ldots, T-1$, and $i, j = 1,\ldots, N$. 
%
We assume that $X_1$ follows the chain's stationary distribution, which is denoted by $\delta \in \bbR^N$, with $i$th component
$\delta_i = \Pr\{X_1 = i\},~ i = 1,\ldots,N.$
%
Recall that a Markov chain's stationary distribution is determined by its probability transition matrix via $\delta = \delta \Gamma$ and $\sum_{i=1}^N \delta_i = 1$.
%
The distribution of an emission $Y_t$ depends only on the corresponding state $X_t$ and no other observations or hidden states: $p\left(y_t|\{X_1,\ldots, X_T\},\{Y_1,\ldots, Y_T\}/ \{Y_t\}\right) = p(y_t|X_t)$.
%
These conditional distributions are governed by state-dependent parameters. If $X_t = i$, then the state-dependent parameter is $\theta^{(i)}$ and we denote the conditional distribution of $Y_t$ given $X_t=i$ by its conditional density or probability mass function, denoted $f^{(i)}(\cdot ; \theta^{(i)})$, or sometimes  $f^{(i)}(\cdot)$.
%
(Fig \ref{fig:HMM}) represents the dependence structure of an HMM.

@@

%
We denote the conditional distribution of $Y_t$ given $Y_{t-1}= y_{t-1}$ and $ X_t=i$ as $f^{(i)}( \cdot | y_{t-1}; \theta^{(i)})$ or simply $f^{(i)}( \cdot | y_{t-1})$.
As an example, we could assume that this conditional distribution is Normal with parameters $\theta^{(i)} = \{\mu^{(i)},\sigma^{(i)},\phi^{(i)}\}$ where:
%
\[
\mathbb{E}(Y_t|Y_{t-1} = y_{t-1},X_t=i) = \phi^{(i)} ~ y_{t-1} ~+ ~(1-\phi^{(i)})  ~\mu^{(i)}
\]
and
\[
\mathbb{V}(Y_t| Y_{t-1} =y_{t-1}, X_t = i) = (\sigma^{(i)})^2.
\]
%
The likelihood for the CarHMM can be easily calculated using the forward algorithm. As previously, let $y$ be the vector of observed emissions. Then
\begin{equation}
    \calL_{\text{CarHMM}}(y;\Theta,\Gamma) = \delta \prod_{t=2}^T \Gamma P(y_t|y_{t-1};\Theta) \mathbf{1}_N
    \label{CarHMM_likelihood}
\end{equation}
where
%
$P(y_t|y_{t-1};\Theta)$ is an $N \times N$ diagonal matrix with $ii$th entry equal to $f^{(i)}(y_t|y_{t-1}; \theta^{(i)})$.
%
\subsection{Continuous-time processes}

While CarHMMs can model auto-correlation within the observation sequence, they fail when observations are taken at irregular time intervals. Another way to model a stochastic process is to use a stochastic differential equation (SDE). As an example, \citep{Michelot:2019} model the movement of an animal as the solution to the following stochastic differential equation:
%
\begin{equation}
    \label{eqn:SDE}
    dY_t = \beta^{(X_t)}(\gamma^{(X_t)} - Y_t)dt + \omega^{(X_t)} dW_t
\end{equation}
%
where $X_t$ follows some stochastic process which defines the hidden behaviour of the animal at time $t$, $\beta^{(X_t)}$ relates to rate at which the process returns to its mean value, $\gamma^{(X_t)}$ is the long-term mean value of the process, $\omega^{(X_t)}$ is related to short-term variance, and $W$ is a Wiener process. This SDE is referred to as a state-switching Ornstein-Uhlenbeck process. Unlike HMMs, the time index $t \in \mathbb{R}$ exists in continuous time and is not necessarily an integer. If the behavioral state $X_t$ is known and does not change between observations, the solution to (eqn \ref{eqn:SDE}) is known to be the following \citep{Michelot:2019}:
%
\begin{align}
    Y_{t+\Delta t} | X_{t} \sim \mathcal{N}\left((1-e^{-\beta^{(X_t)}\Delta t})\gamma^{(X_t)} + e^{-\beta^{(X_t)}\Delta t} Y_t,\quad \frac{\omega^{(X_t)^2}}{2\beta^{(X_t)}} (1-e^{-2\beta^{(X_t)}\Delta t})\right)
    \label{eqn:OU_sol}
\end{align}
%
where $\Delta t$ is the time difference between any two observations $Y_t$ and $Y_{t+\Delta t}$. Most continuous time models are difficult to incorporate into an HHMM and require MCMC methods to fit. However, under certain conditions, the CarHMM is equivalent to a state-switching Ornstein-Uhlenbeck process. This gives new interpretation to the learned parameters of the CarHMM in the context of a continuous-time model.

@@

A \textit{hierarchical hidden Markov model} contains both a coarse-scale process and a fine-scale process, both of which are HMMs. The coarse-scale process is a hidden Markov model as defined previously, where $X_1, \ldots, X_T$ make up an unobserved Markov chain with $N$ possible states and $Y_1,\ldots, Y_T$ are the corresponding observed responses.   
%
In the hierarchical setting, each state $X_t$ emits another sequence of fine-scale unobserved states, $X_t^* \equiv (X_{t,1}^*,\ldots, X_{t,T_t^*})$ and a sequence of fine-scale observed emissions, $Y_t^* \equiv (Y_{t,1}^*,\ldots, Y_{t,T_t^*})$. This fine-scale process can be either instead of or in addition to the coarse-scale observations $Y_t$. The length of each fine-scale process is denoted as $T^*_t$ and can often be controlled by the practitioner. Conditional on the coarse-scale hidden state $X_t$, the joint fine-scale process, ($X_t^*, Y_t^*$), is a hidden Markov model with parameters depending on the value of $X_t$. Specifically, given that $X_t=i$, $X_t^*$ is a Markov chain with states $1,\ldots, N_t^*$. The distribution of $X_t^*$, conditional on $X_t=i$, is given by an $N^*_t \times N^*_t$ transition probability matrix $\Gamma^{*(i)}$ and initial probability, denoted by the $N_t^*$-vector $\delta^{*(i)}$, which we assume is equal to the stationary distribution of the chain. For simplicity, we take $N_t^* \equiv N^*$ although this is not necessary.
%
The distribution of $Y_{t, t^*}$ given $X_{t, t^*}=i^*$ and $X_t=i$ is governed by a parameter $\theta^{(i,i^*)}$ and has density or probability mass function denoted $f^{*(i,i^*)}\left(\cdot; \theta^{(i,i^*)}\right)$ or simply $f^{*(i,i^*)}(\cdot)$. Let $\Theta^{*(i)}=\left(\theta^{(i,1)}, \ldots, \theta^{(i,N^*)}\right)$, the fine-scale emission parameter vector corresponding to $X_t=i$.
%
Given the coarse-scale states, $X_1,\ldots, X_T$, the $T$ fine-scale processes $(X_1^*, Y_1^*), \ldots, (X_T^*, Y_T^*)$, are independent HMMs. Depending upon the process being modeled, it is possible to force certain parameters to be shared across different coarse or fine states. For example, in the killer whale case study in section \ref{sec:case_study}, we force the fine-scale emission parameters to be shared across coarse-scale hidden states (i.e. $\theta^{(1,i^*)} = \ldots = \theta^{(N,i^*)}$ for $i^* = 1, \ldots, N^*$). Figure \ref{fig:HHMM} shows a graphical representation of the dependence structure for an HHMM.

Due to the nested structure of the hierarchical hidden Markov model, the likelihood is easy to calculate via the forward algorithm.
%
Let $y$ be the $T$-vector of the observed coarse-scale emissions and
$y^*$ be the $(T_1^* + \cdots + T_T^*)$-vector of the observed fine-scale emissions.
%
Let $\Theta^*$ denote the collection of all fine-scale emission parameters,
$\Theta^{*(i)}$, $i=1,\ldots, N$, and let $\Gamma^*$ denote the collection of all fine-scale transition probability matrices, $\Gamma^{*(i)}$, $i=1,\ldots, N$.
%
The likelihood of the observed data is
%
\[
\calL_{\text{HHMM}}(y,y^*;\Theta,\Theta^*,\Gamma,\Gamma^*) = \delta P(y_1,y_1^*;\Theta,\Theta^*,\Gamma^*) \prod_{t=2}^T \Gamma P(y_t,y_t^*;\Theta,\Theta^*,\Gamma^*) \mathbf{1}_N
\]
%
where $P(y_t,y_t^*;\Theta,\Theta^*,\Gamma^*)$ is an $N \times N$ diagonal matrix with $ii$th entry corresponding to $X_t=i$ and equal to 
$f^{(i)}(y_t)\calL_{\text{HMM}}\left(y_t^*;\Theta^{*(i)},
\Gamma^{*(i)}\right)$.

@@

The HMM with discrete Fourier transform, or HMM-DFT, incorporates hierarchical structure into an HMM differently than an HHMM. In particular, the fine-scale process is no longer modeled with an HMM and instead summarized using its Fourier transform. For simplicity, we assume that the length of the fine-scale processes is constant (i.e. that $T^*_t = T^*$), although this need not be the case in general. Suppose that the fine-scale process $y^*_t$ does not switch hidden states, but does exhibit significant periodic behaviour. We then suggest using the discrete Fourier transform (DFT) on $y^*_t$:
%
\begin{align*}
    DFT\{y^*_t\}(k) := \hat{y}^{*(k)}_{t} = \sum_{t^* = 1}^{T^*} y^*_{t,t^*}\exp\left(-i \frac{2\pi k}{T^*} (t^*-1)\right) \qquad k = 0, 1, \ldots, T^*-1
\end{align*}
%
Summary statistics can then drastically reduce the dimension of $\hat{y}^*_t$. One such example is as follows:
%
\begin{equation}
    \label{eqn:z}
    z_t^{*(1)} := \mathcal{R}\left(\hat{y}^{(0)}_t\right) \qquad z_t^{*(2)} := \frac{1}{T^*}\sum_{k=1}^{\tilde{\omega}}|\hat{y}^{(k)}_t|^2
\end{equation}
%
$z_t^{*(1)}$ is the average value of $y^*_t$ and $z_t^{*(2)}$ is the squared 2-norm of the component of $y^*_t$ that can be attributed to frequencies between $1$ and $\tilde{\omega}$ periods per window length $T^*$. The maximum frequency $\tilde{\omega}$ is a problem-specific tuning parameter which should be selected with care. These summary statistics are just one possible choice to describe each window; other choices include the dominant frequency and amplitude of $y^*_t$. Figure (\ref{fig:fourier_example}) visually displays the process of transforming $y^*$ into $z^*$.

Once $z^*_t$ is calculated, it can be treated as an observation of the coarse-scale HMM and incorporated into the emission distribution $f^{(i)}\left(y_t,z^*_t;\theta^{(i)}\right)$, or more succinctly $f^{(i)}\left(y_t,z^*_t\right)$. In total, the likelihood of the hierarchical HMM-DFT is as follows:
\begin{equation}
    \calL_{\text{HMM-DFT}}(y,z^*;\Theta,\Gamma) = \delta P(y_1,z^*_1;\Theta) \prod_{t=2}^T \Gamma P(y_t,z^*_t;\Theta) \mathbf{1}_N
    \label{HMMDFT_likelihood}
\end{equation}
where $P(y_t,z^*_t;\Theta)$ is an $N \times N$ diagonal matrix with $ii$th entry equal to $f^{(i)}\left(y_t,z^*_t;\theta^{(i)}\right)$.

### 2020-07-16 Report/main.tex
Updates from Overleaf

Recent advances in high-frequency tagging
technology have made available a vast
amount of movement data in a variety of fields. This
rich new data can exhibit simultaneous behavioural processes occurring at different
time scales, resulting in very complicated dependence structures. One can model these processes
through a hierarchical hidden Markov model
(HHMM), where the system is modelled as a
nested structure of hidden Markov models.
At very short time scales, however, many assumptions of traditional HMMs are violated. We demonstrate how
to incorporate fine-scale processes into the
larger structure of HHMMs while maintaining computational efficiency. We apply our
method to dive and accelerometer data collected from a northern resident killer whale off
the coast of British Columbia, Canada.

>>>>

Recent advances in high-frequency tagging
technology have made a vast
amount of movement data available in a variety of fields. This
rich data can exhibit simultaneous behavioural processes occurring at different
time scales, resulting in very complicated dependence structures. These processes can be modelled 
through a hierarchical hidden Markov model
(HHMM), where the system is modelled as a
nested structure of hidden Markov models (HMMs).
At very short time scales, however, many assumptions of traditional HMMs are violated. We demonstrate how
to incorporate fine-scale processes into the
larger structure of HHMMs while maintaining computational efficiency. We apply our
method to dive and accelerometer data collected from a northern resident killer whale off
the coast of British Columbia, Canada.
